Synthesizing Quotient Lenses

### 2018-03-13 papers/QuotientLenses/main.tex

\newcommand{\OpticianRuntime}{\textbf{OO}}
\newcommand{\SystemOnOptician}{\textbf{QO}}
\newcommand{\SystemOnBenchmarks}{\textbf{QQ}}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

\begin{figure}[t]
\includegraphics{generated-graphs/asts.eps}
\caption{Count of benchmark programs definable using a given AST count. We find
that it takes far fewer AST nodes to define benchmark lenses using QRE
synthesis than with Optician without QRE synthesis or without synthesis.}
\label{fig:asts}
\end{figure}

@@

We are interested in comparing three different approaches, which vary
in the amount of synthesis used. 
In the first approach, which we call \QRESize{} for QRE Synthesis, the programmer uses
QRE synthesis.  She must write QRE specifications of the source and
target formats and she may give examples.
In the second approach, which we call \canonizeAndSpecSize{} for
Bijective Synthesis, the
programmer uses bijective lens synthesis \`a la the original version
of Optician.  She must write canonizers by hand, along with the
regular expressions describing the external
representations of the source and target formats. (The internal
regular expressions can be inferred from the canonizers.) She may also
provide examples to help in the synthesis of the bijective lens.
In the third approach, which we call \LensAndSpecSize{} for No Synthesis, the
programmer writes the lens between the source and target lenses
entirely by hand.

To measure the programmer effort using each of these approaches, we
calculate three different quantities for each problem in the benchmark suite:
%
\begin{itemize}
  \item[\QRESize{}:] 
  The number of AST nodes in the QRE specifications for the source and
  target formats, including examples. 
  \item[\canonizeAndSpecSize{}:] 
  The sum of (1) the number of AST nodes in $W(q)$ for each QRE $q$ in the source and target
  formats, (2) the number of AST nodes $\canonize(q)$ for each such QRE $q$ with a
  non-trivial canonizer, and (3) the number of AST nodes in the
  examples.  We use (1) and (2) to estimate the burden of describing
  the external source and target formats and of writing the requisite canonizers
  by hand.  We count the nodes in the examples because they would be
  fed to the bijective synthesizer.  
  These counts are an approximation, as both $W(q)$ and $\canonize(q)$ are
  automatically generated from the QRE, and it is possible that a human-written
  version might be smaller.
  \item[\LensAndSpecSize{}:] The sum of (1) the number of AST nodes in
  $W(q)$ for each QRE $q$ in the source and target formats and (2) the
  number of AST nodes in the synthesized QRE lens.  We use (1) to
  estimate the burdern for describing the source and target formats
  and (2) to estimate the burdern for writing the appropriate lens by
  hand. These counts are also approximations, as
  $W(q)$ and the synthesized lens may be larger than one written by hand.
\end{itemize}

Figure~\ref{fig:asts} shows the number of benchmarks that can be expressed using
at most a given number of AST nodes. On average, \canonizeAndSpecSize{} used
44\% more AST nodes than \QRESize{}, using an average of 237 more AST nodes. On
average, \LensAndSpecSize{} used 193\% more AST nodes than \QRESize{}, using an
average of 1267 more AST nodes. This suggests that introducing QREs saves
programmers significant effort compared to both pre-QRE Optician and basic
Boomerang.

@@

\begin{figure}[t]
\centering
\begin{subfigure}[b]{.49\textwidth}
\centering
\includegraphics{generated-graphs/times_opt}
\caption{}
\label{subfig:lenssize}
\end{subfigure}
\begin{subfigure}[b]{.49\textwidth}
\includegraphics{generated-graphs/times_new.eps}
\caption{}
\label{subfig:examplesused}
\end{subfigure}
\caption{Runtimes of Optician and our system. In (a), we run Optician and our
  system on Optician's benchmarks. We find that there is only a negligible
  performance overhead incurred by using QREs. In (b), we run Optician with QREs
  on the Optician benchmarks we modified to use QREs. We find that it is able to
  synthesize all quotient lenses in under 20 seconds, and typically finished in
  under 5 seconds.}
\label{fig:times}
\end{figure}

To assess the performance impact of adding QRE support to Optician, we measured
the running time of the following configurations:
\begin{itemize}
  \item[\OpticianRuntime{}] pre-QRE Optician running on its own benchmarks.
  \item[\SystemOnOptician{}] Optician with QREs running on the pre-QRE
    benchmarks.
  \item[\SystemOnBenchmarks{}] Optician with QREs running on the benchmarks that
    were altered from Optician to include quotients.
\end{itemize}

We summarize the results of these experiments in Figure~\ref{fig:times}.  The
new version of Optician was able to synthesize all of pre-QRE Optician's
benchmarks at a speed competitive with the old version.  There is a small amount
of additional overhead introduced by QREs when calculating $W(q_1)$ and $K(q_1)$,
resulting in a very slight decrease in performance.

>>>>

%\newcommand{\QOpt}{Optician_Q}
\newcommand{\QOpt}{QRE-enhanced Optician}
\newcommand{\OpticianRuntime}{\textbf{OO}}
\newcommand{\QREOptician}{\textbf{OQ}}
\newcommand{\SystemOnOptician}{\textbf{QO}}
\newcommand{\SystemOnBenchmarks}{\textbf{QQ}}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

\begin{figure}[t]
\includegraphics{generated-graphs/asts.eps}
\caption{AST node measurements for each of the three approaches
on each of the 10 non-bijective benchmark problems.
QRE Synthesis requires far fewer AST nodes than the other
two approaches. \ksf{Anders, is there an order to the benchmarks in
this list?  It might be useful to list them in order of increasing
complexity of the format, perhaps measured by the sum of the number of
AST nodes in the source and target format descriptions.  It seems
better to have some order than just listing them randomly, even if it
doesn't show much interesting.}}
%%\caption{Count of benchmark programs definable using a given AST count. We find
%%that it takes far fewer AST nodes to define benchmark lenses using QRE
%%synthesis than with Optician without QRE synthesis or without synthesis.}
\label{fig:asts}
\end{figure}

@@

We are interested in comparing three different approaches, which vary
in the amount of synthesis used. 
In the first approach, which we call \QRESize{} for QRE Synthesis, the programmer uses
QRE lens synthesis.  She must write QRE specifications of the source and
target formats and she may give examples.
In the second approach, which we call \canonizeAndSpecSize{} for
Bijective Synthesis, the
programmer uses bijective lens synthesis \`a la the original version
of Optician.  She must write canonizers by hand, along with 
regular expressions to describe the external
representations of the source and target formats. (The internal
formats can be inferred from the canonizers.) She may also
provide examples to help in the synthesis of the bijective lens.
In the third approach, which we call \LensAndSpecSize{} for No Synthesis, the
programmer writes the lens between the source and target formats
entirely by hand, including the descriptions of the source and target
formats.

For each problem in the benchmark suite, we calculate the following
measures as proxies for the level of programmer effort when using each
the three approaches:

%
\begin{itemize}
  \item[\QRESize{}:] 
  The number of AST nodes in the QRE specifications for the source and
  target formats, including examples. 
  \item[\canonizeAndSpecSize{}:] 
  The sum of (1) the number of AST nodes in $W(q)$ for each QRE $q$ in the source and target
  formats, (2) the number of AST nodes in $\canonize(q)$ for each QRE $q$ with a
  non-trivial canonizer, and (3) the number of AST nodes in the
  examples.  We use (1) to estimate the burden of describing
  the external source and target formats and (2) to estimate the
  burdern of writing the requisite canonizers
  by hand.  We count the nodes in the examples because they would be
  fed to the bijective synthesizer.  
  These counts are an approximation, as both $W(q)$ and $\canonize(q)$ are
  automatically generated from the corresponding QRE $q$, and it is
  possible that a human-written version might be smaller.
  \item[\LensAndSpecSize{}:] The sum of (1) the number of AST nodes in
  $W(q)$ for each QRE $q$ in the source and target formats and (2) the
  number of AST nodes in the synthesized QRE lens.  We use (1) to
  estimate the burdern of describing the source and target formats
  and (2) to estimate the burdern of writing the appropriate lens by
  hand. These counts are also approximations, as
  $W(q)$ and the synthesized lens may be larger than one written by hand.
\end{itemize}

\ksf{Anders: are these figures up to date?}
Figure~\ref{fig:asts} shows each of these measures for the 10
non-bijective problems in the benchmark suite.  On
average, \canonizeAndSpecSize{} used~44\% more AST nodes
than \QRESize{}, requiring an average of~237 more AST nodes. On 
average, \LensAndSpecSize{} used~193\% more AST nodes than \QRESize{}, requiring an
average of~1267 more AST nodes. These figures suggest that introducing QREs saves
programmers significant effort compared to both pre-QRE Optician and basic
Boomerang.

@@

\begin{figure}[t]
\centering
\begin{subfigure}[b]{.49\textwidth}
\centering
\includegraphics{generated-graphs/times_opt}
\caption{}
\label{subfig:lenssize}
\end{subfigure}
\begin{subfigure}[b]{.49\textwidth}
\includegraphics{generated-graphs/times_new.eps}
\caption{}
\label{subfig:examplesused}
\end{subfigure}
\caption{Runtimes measurements. In (a), we run the
  original version of Optician (\OpticianRuntime{}) and \QOpt{}
  (\QREOptician{}) on the Optician benchmarks.  We find that there is only a negligible
  performance overhead incurred by using QREs. 
  In (b), we run \QOpt{} on the 10 Optician benchmarks
  previously edited to make them bijective, after removing those edits
  (In other words, we restored them to their original state and then
  ran \QOpt{}). 
  We find that \QOpt{} is able to
  synthesize all quotient lenses in under 20 seconds, and typically finishes in
  under 5 seconds.}
\label{fig:times}
\end{figure}

To assess the performance of QRE synthesis, we are interested in two
different questions.  First, how does the performance of \QOpt{} 
compare to the performance of the original
version of Optician on benchmarks that do not require QREs?  The
answer to this question tells us how much overhead we have introduced
by adopting the more general mechanism.  
Figure~\ref{fig:times}(a) shows that \QOpt{} was able to synthesize
all of the Optician benchmarks at a speed competative with the old
version. 
There is a small amount
of additional overhead introduced by QREs in calculating the $W$ and
$K$ functions, resulting in a slight decrease in performance.

Second, how much time does it take for \QOpt{} to synthesize a QRE
lens when running on a non-bijective benchmark problem?  
Figure~\ref{fig:times}(b) shows the amount of time required to infer a
lens for each of the 10 benchmark programs.  
We find that \QOpt{} is able to synthesize all quotient lenses in
under 20~seconds, and typically finishes in under 5~seconds.
